---
title: "Peak Bloom Prediction Demo"
author: "Ken Koon Wong, Chao-ping Wu, Anirban Bhatthacharrya"
date: "01/01/2022"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```


# library
```{r}
library(tidyverse)
library(lubridate)
library(jsonlite)
library(httr)
library(curl)
library(geofacet)
library(dagitty)
library(ggdag)
library(gridExtra)
library(broom)
```


## Loading the data

The data for each of the three main sites is provided as simple text file in CSV format.
Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site.

The six columns in each data file are

* _location_ a human-readable location identifier (`string`).
* _lat_ (approximate) latitude of the cherry trees (`double`).
* _long_ (approximate) longitude of the cherry trees (`double`).
* _alt_ (approximate) altitude of the cherry trees (`double`).
* _year_ year of the observation (`integer`).
* *bloom_date* date of peak bloom of the cherry trees (ISO 8601 date `string`). The "peak bloom date" may be defined differently for different sites
* *bloom_doy* days since January 1st of the year until peak bloom (`integer`). January 1st corresponds to `1`.

In R, the data files can be read with `read.csv()` and concatenated with the `bind_rows()` function:

```{r}
cherry <- read.csv("data/washingtondc.csv") %>% 
  bind_rows(read.csv("data/liestal.csv")) %>% 
  bind_rows(read.csv("data/kyoto.csv")) %>%
  mutate(bloom_date = ymd(bloom_date))
  
```


# exploring variable on cherry
```{r}
cherry %>% summary()
# we have Location, lat, long, alt, year, bloom_date and bloom day

# this (bloom_day) will be out response/outcome variable, let's see if bloom_day is normally distributed
shapiro.test(cherry$bloom_doy) #W = 0.9873, p-value = 5.751e-08, not normal
# let's visualize this 
hist(cherry$bloom_doy) # left skewed
hist(log(cherry$bloom_doy)) # tried logging but still not normal, 

#?any difference in site, let's visualize
doy_hist <- cherry %>%
  ggplot(.,aes(x=bloom_doy,fill=location)) +
  geom_histogram(alpha = 0.5) + # interesting, looks like it might be site related 
  ggtitle("Day of year (Histogram)") +
  theme_bw()
# let's test shapiro normality w each site
map(.x=cherry$location %>% unique(),.f=~cherry %>% filter(location == !!.x) %>% pull(bloom_doy) %>% shapiro.test()) # much better !!! all 3 sites p >0.05, no need special transition to predict

```



## Visualizing the time series
```{r}
time_series_all <- cherry %>%
  ggplot(.,aes(x=bloom_date,y=bloom_doy,color=location)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_labels = "%b %Y") + # wow, kyoto data goes back to the year 812 and up... let's filter to a lil bit more recent like 1950
  ggtitle("Time series for all available data") +
  theme_bw()

time_series <- cherry %>%
  filter(bloom_date >= mdy("1-1-1950")) %>%
  ggplot(.,aes(x=bloom_date,y=bloom_doy,color=location)) +
  geom_point() +
  geom_line(alpha=0.5) +
  geom_smooth() +
  scale_x_date(date_labels = "%b %Y") + # very interesting, they all seem to follow a cyclical trend
  ggtitle("Time series of Day of Year to Peak Cherry Blossom") +
  theme_bw()
```

#Cherry trees need a full month of chilly weather below 41 degrees to properly blossom when it gets warmer, according to Naoko Abe, author of The Sakura Obsession. If they don't get that chilly weather, they blossom later because "they can't wake up properly," Abe says. https://www.npr.org/2021/04/05/984470981/what-the-cherry-blossom-bloom-can-tell-us-about-climate-change

# let's do a simple causal diagram and see if we can find other data to generate a prediction model
```{r}
library(dagitty)
library(ggdag)

dag <- dagify(
  bloom_doy ~ cherry_type + cold_more_than_30d + location + altitude,
  cold_more_than_30d ~ location + altitude,
  cherry_type ~ location + altitude,
  exposure = "cold_more_than_30d",
  outcome = "bloom_doy"
  
)


set.seed(11)
dag_plot <- ggdag_status(dag, text_col = "black") +
  theme_dag()

paths(dag)
ggdag_dseparated(dag, controlling_for = c("location","altitude","cherry_type","cold_more_than_30d"))
# looks like have to control for all of the above to d-separate

```


# let's see if we can find above data from available dataset provided
```{r}
read_csv("data/meteoswiss.csv") %>% view() #has other location in switzerland's bloom doy
read_csv("data/USA-NPN_individual_phenometrics_data.csv") %>% view() # interestingly US data has different species of cherry blossom and different state too with tmax of each season and percipitation etc, but no info on our variables in causal diagram (30d cold)
# let's do EDA on the above US data
df <- read_csv("data/USA-NPN_individual_phenometrics_data.csv") 

species <- df %>%
  ggplot(.,aes(x=First_Yes_Year,y=First_Yes_DOY,color=Species)) +
  geom_point() +
  geom_line() +
  facet_geo(~State) + # hmmm... looks like species and state may be factor to bloom_doy, difficult to say with limited data, but from sparse data, species does play a role
  ggtitle("Day of Year to Peak Cherry Blossom faceted by States in US and species") +
  theme_bw()

```

# gridextra to plot all 
```{r}
grid.arrange(doy_hist,time_series,time_series_all,dag_plot)
```



# can we get weather data?
```{r}
# api <- Sys.getenv("OPENWEATHER")
# lat <- cherry$lat %>% unique()
# lon <- cherry$long %>% unique()
# start <- mdy_hms("10-1-1990 00:00:00") %>% as.numeric()
# end <- mdy_hms("10-1-2020 00:00:00") %>% as.numeric()
# 
# his <- paste0("http://history.openweathermap.org/data/2.5/history/city?lat=",lat,"&lon=",lon, "&type=daily&start=",start,"&end=",end,"&appid=",api)
# 

# weather <- fread("data/ed651b7dc6bafceebb59d31406244edf.csv") # going to use data.table to read this since it is faster than dplyr


# weather_clean <- weather %>%
#   mutate(date = ymd_hms(dt_iso)) %>%
#   select(date, location=city_name, temp_min) %>%
#   mutate(location = case_when(
#     location == "Washington" ~ "washingtondc",
#     TRUE ~ tolower(location)
#   ),
#   mdy=date(date)) %>%
#   group_by(mdy,location) %>%
#   mutate(mintemp = min(temp_min))

load("data/weather_clean.rda")

winter_add <- weather_clean %>%
  distinct(location,mdy,mintemp) %>%
  mutate(year = year(mdy),
         month = month(mdy)) %>%
  filter(!month %in% c(3:9)) %>%
  mutate(beginwintermonth = case_when(
    month == 10 ~ 1,
    month == 11 ~ 2,
    month == 12 ~ 3,
    month == 1 ~ 4,
    month == 2 ~ 5
  ),
  adjust_year = case_when(
    month %in% c(10:12) ~ year + 1,
    month %in% c(1:2) ~ year
  )) 

winter_add %>%
  ggplot(.,aes(x=mintemp, fill=location)) +
  geom_histogram(position = "dodge") +
  geom_density() +
  facet_wrap(facets = "adjust_year", ncol = 10, nrow = 10) +
  theme_bw() +
  scale_fill_manual(values=c("black", "red", "green"))


```

# combine weather_add to cherry
```{r}
winter_add2 <- winter_add %>%
  mutate(year = adjust_year) %>%
  select(location, mintemp, year,beginwintermonth) 

c_df <- cherry %>%
  filter(year > 1980) %>%
  left_join(winter_add2, by = c("year","location")) %>%
  group_by(location,year) %>%
  mutate(mean_temp = mean(mintemp)) %>%
  select(location,bloom_doy,year,alt,mean_temp) %>%
  distinct_all()



```

# simple linear regression with mean min_temp from October thru Feb
```{r}

model <- lm(bloom_doy ~ location + mean_temp, data = c_df)

summary(model)
sqrt(mean(model$residuals^2))

# interesting, using mean temperature, the rsq is 0.16, rmse is 7.06. not as good as article ping found https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0027439 where rsq is 0.57-0.76, with rmse of 5.6-6.6days

 
model %>% summary()
model %>% tidy() 

paste0("RMSE: ",sqrt(mean(model$residuals^2)))

```


